Reagents: expressing and composing fine-grained concurrency

### 2011-11-07 paper/paper.tex
finalversion

\paragraph{Expressive}  
Reagents provide a basic set of building blocks for writing concurrent data
structures and synchronizers.  The building blocks include isolated atomic
updates to shared state, and interactive synchronous communication through
message passing.  This blend of isolation and interaction is essential for
expressing the full range of fine-grained concurrent
algorithms~(\secref{choice}).  The building blocks also bake in many common
concurrency patterns, like optimistic retry loops, backoff schemes, and blocking
and signalling.  Using reagents, it is possible to express sophisticated
concurrency algorithms at a higher level of abstraction, while retaining the
performance and scalability of direct, hand-written implementations.

@@

We test this principle empirically in~\secref{performance} by comparing multiple
reagent-based collections to their hand-written counterparts, as well as to
lock-based and STM-based implementations.  The benchmarks include both single
operations used in isolation (where little overhead for reagents is expected)
and combined atomic transfers.  We compare at both high and low levels of
contention.  Reagents perform universally better than the lock- and STM-based
implementations, and are indeed competetive with hand-written lock-free
implementations.

@@

Both \lstinline{push} and \lstinline{tryPop} are \emph{total} operations: they
can in principle succeed no patter what state the stack is in, and fail (and
retry) only due to active interference from concurrent threads.  A true
\lstinline{pop} operation, on the other hand, is \emph{partial}: it is undefined
when the stack is empty.  Often this is taken to mean that the operation should
block until another thread moves the stack into a state on which the operation
is defined.  Partial operations introduce considerable complexity, because
\emph{all} of the operations on the data structure must potentially signal
blocked threads, depending on the changes being performed.  In some cases, it is
possible to cleverly treat signalling in essentially the same way as atomic
updates~\cite{Scherer2004}.

Synchronization and signalling is also subject to cache and memory bandwidth
concerns, but it would take us too far afield to discuss these in depth.
Mellor-Crummey and Scott pioneered the now-common techniques for grappling with
these concerns~\cite{Mellor-Crummey1991}, and we apply their techniques in
implementing the blocking and signaling protocols for reagents.

@@

While \lstinline{tryPop} handles both empty and nonempty stacks, we
can write a variant that drops the empty case:
\begin{lstlisting}
val pop: Reagent[Unit, A] = 
  upd(head) { case (x::xs, ()) => (xs, x) }
\end{lstlisting}
Now our update function is partial.  An invocation \lstinline{s.pop ! ()} will
block the calling thread unless or until the stack is nonempty.  Blocking and
signalling are entirely handled by the reagent implementation; there is
therefore no risk of a lost wakeup.  %TODO: add more here?

@@

Reagents interact through \emph{synchronous swap channels}, which consist of two
complementary \emph{endpoints}.  The method \lstinline{mkChan[A,B]} returns a pair
of type \lstinline{(Endpoint[A,B], Endpoint[B,A])}.
The combinator for communication is \lstinline{swap} (see
\figref{combinators}), which lifts an \lstinline{Endpoint[A,B]} to a
\lstinline{Reagent[A,B]}.  When two reagents communicate on opposite
endpoints, they provide messages of complementary type (\lstinline{A}
and \lstinline{B}, for example) and receive each other's value.  We
call a successful communication a \emph{reaction} between reagents.
On the other hand, if no complementary message is available,
\lstinline{swap} will block until a reaction can take place.  Once
again, the waiting and signalling protocol is managed internally by
the reagent implementation, so lost wakeups are impossible.

@@

Message passing between reagents is synchronous, meaning that both reagents take
part in a single, common reaction.  In the implementation, this works by one
reagent placing an \emph{offer} to react in a location visible to the other.
The reagent accepting the offer combines the accumulated reactions of both
reagents, and attempts to commit them together.  The reagent making the offer
either spinwaits or blocks until the offer is fulfilled; if it spinwaits, it may
later decide to withdraw the offer.  Fulfilling the offer means, in particular,
providing a final ``answer'' value that should be returned by the reagent that
made the offer.  Each offer includes a status field, which is either
\lstinline{Pending}, \lstinline{Rescinded}, or a final answer.  Hence, the
\lstinline{Offer} class is parameterized by the answer type; a
\lstinline{Reagent[A,B]} will use \lstinline{Offer[B]}.  When fulfilling an
offer, a reagent CASes its status from \lstinline{Pending} to the desired final
answer.

@@

For our prototype implementation, we have opted to use an extremely simple,
blocking implementation that replaces each location to be CASed with a sentinel
value, essentially locking the location.  As the \lstinline{Reaction} object is
assembled, locations are kept in a globally-consistent order, which avoids
deadlock.  The advantage of this implementation, other than its simplicity, is
that is has no impact on the performance of single-word CASes to references,
which we expect to be the common case.  Our experimental results
in~\secref{performance} indicate that the blocking implementation provides good
performance---much better than STM or coarse-grained locking---but we expect
that a more sophisticated $k$CAS would further improve the results.

@@

>>>>

\paragraph{Expressive}  
Reagents provide a basic set of building blocks for writing concurrent data
structures and synchronizers.  The building blocks include isolated atomic
updates to shared state, and interactive synchronous communication through
message passing.  This blend of isolation and interaction is essential for
expressing the full range of fine-grained concurrent
algorithms~(\secref{choice}).  The building blocks also bake in many common
concurrency patterns, like optimistic retry loops, backoff schemes, and blocking
and signaling.  Using reagents, it is possible to express sophisticated
concurrency algorithms at a higher level of abstraction, while retaining the
performance and scalability of direct, hand-written implementations.

@@

We test this principle empirically in~\secref{performance} by comparing multiple
reagent-based collections to their hand-written counterparts, as well as to
lock-based and STM-based implementations.  The benchmarks include both single
operations used in isolation (where little overhead for reagents is expected)
and combined atomic transfers.  We compare at both high and low levels of
contention.  Reagents perform universally better than the lock- and STM-based
implementations, and are indeed competitive with hand-written lock-free
implementations.

@@

Both \lstinline{push} and \lstinline{tryPop} are \emph{total} operations: they
can in principle succeed no patter what state the stack is in, and fail (and
retry) only due to active interference from concurrent threads.  A true
\lstinline{pop} operation, on the other hand, is \emph{partial}: it is undefined
when the stack is empty.  Often this is taken to mean that the operation should
block until another thread moves the stack into a state on which the operation
is defined.  Partial operations introduce considerable complexity, because
\emph{all} of the operations on the data structure must potentially signal
blocked threads, depending on the changes being performed.  In some cases, it is
possible to cleverly treat signaling in essentially the same way as atomic
updates~\cite{Scherer2004}.

Synchronization and signaling is also subject to cache and memory bandwidth
concerns, but it would take us too far afield to discuss these in depth.
Mellor-Crummey and Scott pioneered the now-common techniques for grappling with
these concerns~\cite{Mellor-Crummey1991}, and we apply their techniques in
implementing the blocking and signaling protocols for reagents.

@@

While \lstinline{tryPop} handles both empty and nonempty stacks, we
can write a variant that drops the empty case:
\begin{lstlisting}
val pop: Reagent[Unit, A] = 
  upd(head) { case (x::xs, ()) => (xs, x) }
\end{lstlisting}
Now our update function is partial.  An invocation \lstinline{s.pop ! ()} will
block the calling thread unless or until the stack is nonempty.  Blocking and
signaling are entirely handled by the reagent implementation; there is
therefore no risk of a lost wakeup.  %TODO: add more here?

@@

Reagents interact through \emph{synchronous swap channels}, which consist of two
complementary \emph{endpoints}.  The method \lstinline{mkChan[A,B]} returns a pair
of type \lstinline{(Endpoint[A,B], Endpoint[B,A])}.
The combinator for communication is \lstinline{swap} (see
\figref{combinators}), which lifts an \lstinline{Endpoint[A,B]} to a
\lstinline{Reagent[A,B]}.  When two reagents communicate on opposite
endpoints, they provide messages of complementary type (\lstinline{A}
and \lstinline{B}, for example) and receive each other's value.  We
call a successful communication a \emph{reaction} between reagents.
On the other hand, if no complementary message is available,
\lstinline{swap} will block until a reaction can take place.  Once
again, the waiting and signaling protocol is managed internally by
the reagent implementation, so lost wakeups are impossible.

@@

Message passing between reagents is synchronous, meaning that both reagents take
part in a single, common reaction.  In the implementation, this works by one
reagent placing an \emph{offer} to react in a location visible to the other.
The reagent making the offer either spinwaits or blocks until the offer is
fulfilled; if it spinwaits, it may later decide to withdraw the offer.  The
reagent accepting the offer combines the accumulated reactions of both reagents,
and attempts to commit them together. Fulfilling the offer means, in particular,
providing a final ``answer'' value that should be returned by the reagent that
made the offer.  Each offer includes a status field, which is either
\lstinline{Pending}, \lstinline{Rescinded}, or a final answer.  Hence, the
\lstinline{Offer} class is parameterized by the answer type; a
\lstinline{Reagent[A,B]} will use \lstinline{Offer[B]}.  When fulfilling an
offer, a reagent CASes its status from \lstinline{Pending} to the desired final
answer.

@@

For our prototype implementation, we have opted to use an extremely simple,
blocking implementation that replaces each location to be CASed with a sentinel
value, essentially locking the location.  As the \lstinline{Reaction} object is
assembled, locations are kept in a globally-consistent order, which avoids
deadlock.  The advantage of this implementation, other than its simplicity, is
that is has no impact on the performance of single-word CASes to references,
which we expect to be the common case.  Our experimental results
in~\secref{performance} indicate that the blocking implementation provides good
performance---much better than STM or coarse-grained locking---but a more
sophisticated $k$CAS would likely do even better.

@@

The code sketch we have given for channels glosses over several details of the
real implementation, including avoiding fulfilling one's own offer or a single
offer multiple times, and allowing multiple interactions between two reagents
within a single reaction.
